Location and Scale-Invariant Power Transformations for Transforming Data to

### 2025-10-24 manuscript_rmarkdown/manuscript.Rmd
Update section 3.2 and Figure 5

To develop a test for central normality we need to consider two
parameters: the central portion $\kappa$ and outlier rate $\zeta$ as fixed parameters,
and test statistic $\tau_{\text{ecn}}$. We will first define the central portion
$\kappa$.

For each number of samples $n \in \left\{\lfloor 10^\nu \rfloor | \nu \in \left\{0.7500, 0.8125, \ldots, 4.0000 \right \} \right\}$
we randomly drew $m_d = 30000$ ordered sequences $\mathbf{X}$ from $\mathcal{N}(0,1)$.
This dataset was used to compute the critical test statistic values for the central normality test.
Additionally, outliers were inserted by randomly replacing each element of an ordered sequence with a probability $\zeta$.
We then computed the critical test statistic values for the empirical central normality test.
For each sequence we then computed $\tau_{n, \kappa, \zeta}$ for 
$\kappa \in \left\{0.60, 0.70, 0.80, 0.90, 0.95, 1.00\right\}$ and
$\zeta \in \left\{0.0, 0.01, 0.02, 0.05, 0.10, 0.15, 0.20 \right\}$.

@@

Figure \ref{fig:empirical-central-normality-kappa} shows $\tau_{\alpha = 0.05, n, \kappa, \zeta}$
as a function of $n$ for different values of $\kappa$ and $\zeta$. With decreasing
$\kappa$, the test statistic curve decreases. The curves
for $\kappa = 0.60$, $\kappa = 0.70$ and $\kappa = 0.80$ are similar, whereas curves for 
$\kappa \geq 0.90$ are affected by outliers, as can be observed by comparing curves
of the central normality test with those of the empirical variant.

Since the empirical central normality tests assesses whether the central part of a sequence is normally distributed,
we used $\kappa = 0.80$. Critical statistic values for $\kappa = 0.80$ and $\zeta = 0.10$ are shown in Table 
\ref{tab:empirical-central-normality-critical-statistic}.

```{r empirical-central-normality-kappa, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=2.5}
cap <- paste0(
  "Critical test statistic $\\tau_{\\alpha = 0.05, n, \\kappa, \\zeta}$ of the empirical ",
  "central normality test as function of $n$ for several values of central portion $\\kappa$ ",
  "and outlier rates $\\zeta$. ",
  "emp: empirical"
)

@@

To assess type I error rates for the (empirical) central normality test and compare
these to the Shapiro-Wilk test for normality, we randomly drew another 1000 sequences from a
standard normal distribution ($\mathcal{N}(0,1)$ for each 
$n \in \left\{\lfloor 10^\nu \rfloor | \nu \in \left\{0.7500, 0.8125, \ldots, 3.0000 \right \} \right\}$.
A second dataset was created by replacing elements with randomly drawn outliers, as described above.
For each sequence, the p-value for the respective test was used to reject the null
hypothesis that the sequence is (centrally) normal. The null hypothesis was rejected if $p \leq 0.05$.
The type I error rate was determined by computing the fraction of centrally normal sequences rejected this way.
Figure \ref{fig:empirical-central-normality-error_rate} shows that the empirical central normality test had the expected type I error rate for
centrally normal data (with tails of the distribution not being normally distributed),
whereas the central normality test and Shapiro-Wilk test overestimated type I error rates.

```{r empirical-central-normality-error_rate, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=2.5}
cap <- paste0(
  "Observed type I error rates for (central) normality tests in the presence of data derived from normal distributions with and without outliers. ",
  "The empirical central normality (ECN) test uses $\\kappa = 0.80$. CN: central normality"
)

@@

To determine the weighting function parameters for each of the nine combinations,
$m_d=500$ sequences $\mathbf{X}_i$ ($i \in \{1, 2, \ldots, m_d\}$) were randomly 
drawn from randomly parametrised asymmetric generalised normal distributions.
Each distribution was parametrised with a random skewness parameter $\alpha \sim U\left(0.01, 0.99\right)$ 
and shape parameter $\beta \sim U\left(1.00, 5.00 \right)$.
Location and scale parameters were set as $\mu = 0$ and $\sigma = 1$, respectively.
To form each sequence $\mathbf{X}_i$, $n = \lceil 10^\gamma \rceil$ instances were then randomly drawn, 
with $\gamma \sim U\left(\log_{10}50, 3\right)$, resulting in a set of sequences with between $50$ and $1000$ elements each.

Outlier values were then drawn to randomly replace 10 percent of the elements of $\mathbf{X}_i$.
Outlier values were set according to @Tukey1977-xm, as follows.
Let $x^{*} \sim U\left(-2, 2\right)$.
Then the corresponding outlier value was:

\begin{equation}
x_{out} =
\begin{cases}
Q_1 - \left(1.5 - x^{*} \right) \text{IQR} & \text{if } x^{*} < 0 \\
Q_3 + \left(1.5 + x^{*} \right) \text{IQR} & \text{if } x^{*} \geq 0
\end{cases}
\end{equation}

$Q_1$, $Q_3$ and $\text{IQR}$ are the first quartile, third quartile and interquartile range of $\mathbf{X}_i$, respectively.
Outlier values randomly replaced elements in $\mathbf{X}_i$.

>>>>

The critical test statistic for the empirical central normality test depends on 
the significance level $\alpha$, sequence length $n$, central portion $\kappa$
and estimated outlier rate $\zeta$, and was set through simulation. 
For each number of samples $n \in \left\{\lfloor 10^\nu \rfloor | \nu \in \left\{0.7500, 0.8125, \ldots, 4.0000 \right \} \right\}$
we randomly sampled $m_d = 30000$ ordered sequences $\mathbf{X}$ from $\mathcal{N}(0,1)$.
Outliers were inserted by randomly replacing elements of an ordered sequence with probability 
$\zeta \in \left\{0.0, 0.01, 0.02, 0.05, 0.10, 0.15, 0.20 \right\}$. 
Outlier values were set according to @Tukey1977-xm, as follows.
Let $x^{*} \sim U\left(-2, 2\right)$.
Then the corresponding outlier value was:

\begin{equation}
\label{eqn:outlier_rate}
x_{out} =
\begin{cases}
Q_1 - \left(1.5 - x^{*} \right) \text{IQR} & \text{if } x^{*} < 0 \\
Q_3 + \left(1.5 + x^{*} \right) \text{IQR} & \text{if } x^{*} \geq 0
\end{cases}
\end{equation}

Here, $Q_1$, $Q_3$ and $\text{IQR}$ are the first quartile, third quartile and interquartile range of $\mathbf{X}_i$, respectively.

For each sequence we then computed $\tau_{\kappa}$ for $\kappa \in \left\{0.60, 0.70, 0.80, 0.90, 0.95, 1.00\right\}$.

@@

Figure \ref{fig:empirical-central-normality-kappa} shows the critical test static
$\tau_{\alpha = 0.05, n, \kappa, \zeta}$
as a function of $n$ for different values of $\kappa$ and $\zeta$.
With decreasing $\kappa$, the test statistic curve decreases.
The curves for $\kappa = 0.60$, $\kappa = 0.70$ and $\kappa = 0.80$ are similar,
whereas curves for $\kappa \geq 0.90$ were affected by outliers.

We use $\kappa = 0.80$ and $\zeta = 0.10$ for the empirical central normality tests 
in the remainder of the manuscript, unless noted otherwise.
Critical statistic values of the empirical central normality test for 
$\kappa = 0.80$ and $\zeta = 0.10$ are shown in Table 
\ref{tab:empirical-central-normality-critical-statistic}.

```{r empirical-central-normality-kappa, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=2.5}
cap <- paste0(
  "Critical test statistic $\\tau_{\\alpha = 0.05, n, \\kappa, \\zeta}$ of the empirical ",
  "central normality test as function of $n$ for several values of central portion $\\kappa$ ",
  "and estimated outlier rates $\\zeta$. ",
  "emp: empirical"
)

@@

To assess type I error rates for the empirical central normality test 
($\kappa = 0.80$ and $\zeta \in \left\{0.00, 0.05, 0.10\right\}$) and compare
these to the Shapiro-Wilk test for normality, we randomly drew another 10000 sequences from a
standard normal distribution ($\mathcal{N}(0,1)$ for each 
$n \in \left\{\lfloor 10^\nu \rfloor | \nu \in \left\{0.7500, 0.8125, \ldots, 3.0000 \right \} \right\}$.
Each sequence was additionally randomly populated with outliers at 5% and 10% rates, as earlier.

For each sequence, the p-value for the respective test was used to reject the null
hypothesis that the sequence is (centrally) normal. The null hypothesis was rejected if $p \leq 0.05$.
The type I error rate was determined by computing the fraction of centrally normal sequences rejected this way.
Figure \ref{fig:empirical-central-normality-error_rate} shows that when the estimated outlier rate 
for the empirical central normality test matches the outlier rate in the data,
the test yields an accurate Type I error rate.
The Shapiro-Wilk test is overly conservative in the presence of any outliers. 
Empirical central normality tests where the estimated outlier rate is lower
than the actual outlier rate are likewise conservative, but to a lesser degree. 
Empirical central normality tests where the estimated outlier rate is higher 
than the actual outlier rate are optimistic.

```{r empirical-central-normality-error_rate, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=2.5}
cap <- paste0(
  "Observed type I error rates for central normality tests in the presence of data sampled from normal distributions ",
  "without outliers, or with a 5% or 10% outlier rate. ",
  "The empirical central normality (ECN) test used $\\kappa = 0.80$, and was further parameterised by the ",
  "estimated outlier rate $\\zeta = 0.00$ (ECN (0 %)), $\\zeta = 0.05$ (ECN (5 %)) and $\\zeta = 0.10$ (ECN (10%))."  
)

@@

To determine the weighting function parameters for each of the nine combinations,
$m_d=500$ sequences $\mathbf{X}_i$ ($i \in \{1, 2, \ldots, m_d\}$) were randomly 
drawn from randomly parametrised asymmetric generalised normal distributions.
Each distribution was parametrised with a random skewness parameter $\alpha \sim U\left(0.01, 0.99\right)$ 
and shape parameter $\beta \sim U\left(1.00, 5.00 \right)$.
Location and scale parameters were set as $\mu = 0$ and $\sigma = 1$, respectively.
To form each sequence $\mathbf{X}_i$, $n = \lceil 10^\gamma \rceil$ instances were then randomly drawn, 
with $\gamma \sim U\left(\log_{10}50, 3\right)$, resulting in a set of sequences with between $50$ and $1000$ elements each.
Outlier values were then drawn to randomly replace 10 percent of the elements of $\mathbf{X}_i$.
